---
title: "BLS Data Project Case Study"
author: "Greg York"
date: "`r Sys.Date()`"
output: html_document
---
## Case Study Introduction  

For my capstone project for the Google Certificate in Data Analytics on Coursera,
I wanted to try something a bit different than the provided business cases. 
I've always been interested in policy, and the use of data to inform our
understanding of people and societal trends.  

One of the public data sets the course recommended was from the 
[BLS](https://www.bls.gov/cps/tables.htm). It includes historical data
concerning labor force statistics, with some data reaching back as far as 1936.  

This case study is more of a data exploration. I am getting practice in locating,
downloading, cleaning, organizing, loading, transforming, plotting, and 
publishing using a variety of tools, including R (and RStudio). I do not have
a preconceived question in my mind before I start, but I want to see where
the data takes me.  

### The Setup  

First thing, I have to load the packages I'll need for the analysis. So far,
I have been leaning heavily on the package 'tidyverse'. Dealing with all of this
Excel data, I have been using 'readxl' to (some) success.

```{r Library, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
```

Some of the data on the BLS website is not in spreadsheet format. The text files
included are not delineated in any way (they are designed to be human-readable)
so importing that data using base R or the library 'readr' proved impossible 
(for me). So that seemed like a dead-end.  

The PDFs were also a challenge. The data looks like it's in a tabular form.
The library 'tabulizer' would be useful, but it has not been updated in quite awhile and would not load in my version of R. Instead, I used a utility called
[Tabula](https://tabula.technology/) which managed to pull most of what I needed
from the PDFs into a .csv format so I could clean it up in Excel or R to use.

## First Data Set, First Plot  

So just to get my feet wet, I went ahead and looked at the first table on the 
BLS website, on Employment Status. This XLSX file has general employment trends
reaching back to 1951.

```{r plot_01}
file <- "bls_project/1950-2021_general_employment_status_edited.xlsx"

employment_df <- read_excel(file)

## First Plot Code ------------------
employment_df <- employment_df[1:71,]

ggplot(data = employment_df, aes(x = Year)) +
        geom_line(aes(y = labor_force_percent_of_population,
                      color = 'labor_force_percent_of_population',
                      group = 1), lwd = 2) +
        geom_line(aes(y = employed_percent_of_population,
                      color = 'employed_percent_of_population',
                      group = 1), lwd = 2) +
        geom_vline(xintercept = 2000, color = "purple", linetype = "longdash") +
        labs(y = "Percent of Population") +
        scale_color_discrete(name = "Historical Employment 1951-2021",
                labels=c("Employed", "In Labor Force")) +
        geom_text(mapping=aes(x=2000, y=67.4, label = "Year 2000")) +
        theme(axis.text.x = element_text(angle = 45),
              legend.position = c(.1,.9))
```


```{r cars}
summary(cars)
```

```{r pressure, echo=FALSE}
plot(pressure)
```
